The function--sheaf dictionary for smooth commutative group schemes over finite fields, with applications over local fields

### 2014-08-20 dictionary.tex
Finished Applications.
Still needed: a complete re-write of the Introduction.

@@

Now suppose $X_K$ is a connected commutative algebraic group over $K$ that admits a Néron model $X$ over $R$.

A {\it quasicharacter of $X_K(K)$} is a group homomorphism $X_K(K) \to \EEx$ that factors through $X_K(X) = X(R)\to X(R_n)$, for some $n\in\NN_0$;
the group of quasicharacters of $X_K(K)$ will be denoted by $\Hom_\text{qc}(X_K(K),\EEx)$.

Then $\Gr_R(X)$ is commutative group scheme over $\Fq$
with component group scheme
\[
\pi_0(\Gr_R(X))= \pi_0(X) \times_S S_0
\]
and $\Fq$-rational points
\[
\Gr_R(X)(\Fq)\iso X_K(K),
\]
canonically.
For every $n\in \NN_0$,
$\Gr^R_n(X)$ is {\it smooth} commutative group scheme over $\Fq$
with component group scheme
$\pi_0(\Gr^R_n(X))= \pi_0(X) \times_S S_0$.

For every non-negative integer $n$, the morphism of $\Fq$-schemes $P^R_n(X) : \Gr_R(X) \to \Gr^R_n(N)$ induces a functor
\[
P^R_n(X)^* : \QC(\Gr^R_n(X)) \to \mathcal{C}(\Gr_R(X)),
\]
as in Proposition~\ref{prop:pullback}.
This functor is full and faithful, so we may view $\QC(\Gr^R_n(X))$ as a full subcategory of $\mathcal{C}(\Gr_R(X))$.

@@

Note that we may informally regard $\QC(\Gr_R(X))$ a `union' of the full subcategories $\QC(\Gr^R_n(X))$ in $\mathcal{C}(\Gr_R(X))$.

\begin{theorem}\label{thm:QCX}
Let $X_K$ be a commutative algebraic group over $K$ with Néron model $X$ over $R$. 
\begin{enumerate}
\item[(1)] There is a canonical short exact sequence of abelian groups 
\[
0 \to \Hh^2(\pi_0(X)_{\bFq},\EEx)^{\Frob{}} \to \QCiso{\Gr_R(X)} \to \Hom_\text{qc}(X_K(K),\EEx) \to 0.
\] 
\item[(2)] For all quasicharacter sheaves $\qcs{F}$, $\qcs{F}'$ on $\Gr_R(X)$, and for every $\rho \in \Hom(\qcs{F},\qcs{F}')$, either $\rho$ is trivial or $\rho$ is an isomorphism.
\item[(3)] For all quasicharacter sheaves $\qcs{F}$ on $\Gr_R(X)$,
\[
\Aut(\qcs{F}) \iso \Hom((\pi_0(X)_{\Fq})_{\Weil{}},\EEx),
\]
canonically.
\end{enumerate}
\end{theorem}

@@

@@

Let $K$ and $K'$ be local fields with rings of integers $R$ and $R'$ and residue fields $k$ and $k'$, respectively. Set $S = \Spec(S)$ and $S_n = \Spec(R_n)$; likewise define $S' = \Spec{R'}$ and $S'_{n'} = \Spec{R'_{n'}}$.

Let $X_K$ and $X'_{K'}$ be connected commutative algebraic groups over $K$ and $K'$, respectively, that admit Néron models $X$ and $X'$ over $R$ and $R'$, respectively.
Suppose $m$ and $m'$ are non-negative integers such that 
\begin{equation}\label{eq:schematic_transfer}
R_m \iso R'_{m'}
\qquad
\text{and}
\qquad
X \times_{S} S_m \iso X' \times_{S'} S'_{m'}.
\end{equation}
as smooth group schemes over $R_m$. 
Then
$
\Gr^R_m(X) \iso \Gr^{R'}_{m'}(X')
$
as smooth group schemes over $\Fq$ and, by Proposition~\ref{prop:pullback} the isomorphism above determines an equivalence of categories
\begin{equation}\label{eq:categorical_transfer}
\QC(\Gr^R_m(X)) \cong \QC(\Gr^{R'}_{m'}(X'))
\end{equation}
which induces a group isomorphism
$
\QCiso{\Gr^R_m(X)} \cong \QCiso{\Gr^{R'}_{m'}(X')}
$
and thence, by trace of Frobenius, an isomorphism
\begin{equation}\label{eq:transfer}
\Hom(X(R_m),\EEx) \iso \Hom(X'(R'_{m'}),\EEx).
\end{equation}
Recall that $\Hom(X(R_m),\EEx)$ is a subgroup of $\Hom_\text{qc}(X_K(K),\EEx)$ and $\Hom(X'(R'_{m'}),\EEx)$ is a subgroup of $\Hom_\text{qc}(X_K(K),\EEx)$.
The isomorphism \eqref{eq:transfer} is an instance of {\it transfer} of certain quasicharacters between $X_K(K)$ and $X_{K'}(K')$.

Such a relationship can exist between quasicharacters of algebraic tori over local fields, even when the characteristics of $K$ and $K'$ differ
\citelist{\cite{chai-yu:01a} \cite{deligne:84a}*{\S 1}}.

Suppose $T$ and $T'$ are tori over non-archimedian local fields $K$ and $K'$,
splitting over $L$ and $L'$ respectively.
Then $T$ and $T'$ are said to be \emph{$N$-congruent} \cite{chai-yu:01a}*{\S 2} if there are isomorphisms
 \begin{align*}
  \alpha : \OL/\pi_K^N \OL &\to \OO{L'}/\pi_{K'}^N \OO{L'} \\
  \beta : \Gal(L/K) &\to \Gal(L'/K') \\
  \phi : X^*(T) &\to X^*(T')
 \end{align*}
 satisfying the conditions
 \begin{enumerate}
  \item $\alpha$ induces an isomorphism $\OK/\pi_K^N \OK \to \OO{K'}/\pi_{K'}^N \OO{K'}$,
  \item $\alpha$ is $\Gal(L/K)$-equivariant relative to $\beta$, and
  \item $\phi$ is $\Gal(L/K)$-equivariant relative to $\beta$.
 \end{enumerate}
If $T$ and $T'$ are $N$-congruent then $\alpha$, $\beta$ and $\phi$ determine an isomorphism
\begin{equation}\label{transfer}
  \Hom_N(T(K), \EEx) \iso \Hom_N(T'(K'),\EEx).
\end{equation}
Note that if $T$ and $T'$ are $N$-congruent, then they are $N'$ congruent for every
$N' \leq N$, the isomorphism \eqref{transfer} is depth-preserving at integers less than $N$.

One of the main results of \cite{chai-yu:01a} gives an isomorphism of group schemes between the
$\TT_n$ and $\TT'_n$ assuming that $T$ and $T'$ are sufficiently congruent.
They define a quantity $h$ as the smallest integer so that $\pi^h$ lies in the
Jacobian ideal associated to a natural embedding of $T$ into an induced torus \cite{chai-yu:01a}*{\S 8.1}.
Then it follows from \cite{chai-yu:01a}*{Thm. 8.5} that if $N > 3h$ and $T$ and $T'$ are $N$-congruent
then there is a canonical isomorphism of smooth group schemes
 $
  \TT_{N-3h} \to \TT'_{N-3h}
 $
 determined by $\alpha, \beta$ and $\phi$.
Combining this with Theorem~\ref{thm:application} gives the geometrization and categorification of the transfer of characters.

\begin{proposition}\label{prop:transfer}
 With notation above, suppose that $T$ and $T'$ are $N$-congruent and $N > 3h$.  Set $n \ceq N-3h$.
 Then there is a canonical equivalence of categories
 \[
  \QC(\GN{T}'_{n}) \to \QC(\GN{T}_{n})
 \]
 determined by $\alpha, \beta$ and $\phi$ inducing the isomorphism
 \[
  \Hom_{<n}(T'(K'), \EEx) \to \Hom_{<n}(T(K), \EEx)
 \]
 through the trace of Frobenius.
\end{proposition}

\subsection{Algebraic tori and Abelian varieties over local fields}

\begin{corollary}\label{cor:application}
For every algebraic torus $T$ over $K$, the trace of Frobenius
  \[
\trFrob{T}:  \QCiso{\GN{T}} \to \Hom_{\text{qc}}(T(K),\EEx)
  \]
is an isomomorphism.
Moreover, the isomorphism $\trFrob{T}$ is a natural transformation
between the additive functors
\begin{enumerate}
\item[$F_{\QC}$:] $T \mapsto \QCiso{\GN{T}}$
\item[$F_{\text{qc}}$:] $T \mapsto \Hom_{\text{qc}}(T(K),\EEx)$
\end{enumerate}
from the additive category of algebraic tori over $K$ to the category of abelian groups.
\end{corollary}

>>>>

Suppose $X_K$ is a connected commutative algebraic group over $K$ that admits a Néron model $X$ over $R$.
%
A {\it quasicharacter of $X_K(K)$} is a group homomorphism $X_K(K) \to \EEx$ that factors through $X_K(X) = X(R)\to X(R_n)$, for some $n\in\NN_0$;
the group of quasicharacters of $X_K(K)$ will be denoted by $\Hom_\text{qc}(X_K(K),\EEx)$ and the subgroup of those that factor through $X(R_n)$ will be denoted by $\Hom_n(X_K(K),\EEx)$.
%
In this section we explain how to geometrize and categorify quasicharacters of $X_K(K)$ using quasicharacter sheaves.

@@

The Greenberg transform $\Gr_R(X)$ of $X$ is commutative group scheme over $\Fq$ with component group scheme
\[
\pi_0(\Gr_R(X))= \pi_0(X) \times_S S_0
\]
and $\Fq$-rational points
\[
\Gr_R(X)(\Fq)\iso X_K(K),
\]
canonically.
For every $n\in \NN_0$,
$\Gr^R_n(X)$ is {\it smooth} commutative group scheme over $\Fq$
with component group scheme
$\pi_0(\Gr^R_n(X))= \pi_0(X) \times_S S_0$.
%
For every non-negative integer $n$, the morphism of $\Fq$-schemes $P^R_n(X) : \Gr_R(X) \to \Gr^R_n(N)$ induces a functor
\[
P^R_n(X)^* : \QC(\Gr^R_n(X)) \to \mathcal{C}(\Gr_R(X)),
\]
as in Proposition~\ref{prop:pullback}.
This functor is full and faithful, so we may view $\QC(\Gr^R_n(X))$ as a full subcategory of $\mathcal{C}(\Gr_R(X))$.

@@

%Note that we may informally regard $\QC(\Gr_R(X))$ a `union' of the full subcategories $\QC(\Gr^R_n(X))$ in $\mathcal{C}(\Gr_R(X))$.

\begin{theorem}\label{thm:QCXK}
Let $X_K$ be a commutative algebraic group over $K$ with Néron model $X$ over $R$. 
\begin{enumerate}
\item[(1)] There is a canonical short exact sequence of abelian groups 
\[
0 \to \Hh^2(\pi_0(X)_{\bFq},\EEx)^{\Frob{}} \to \QCiso{\Gr_R(X)} \to \Hom_\text{qc}(X_K(K),\EEx) \to 0.
\] 
\item[(2)] For all quasicharacter sheaves $\qcs{F}$, $\qcs{F}'$ on $\Gr_R(X)$, and for every $\rho \in \Hom(\qcs{F},\qcs{F}')$, either $\rho$ is trivial or $\rho$ is an isomorphism.
\item[(3)] For all quasicharacter sheaves $\qcs{F}$ on $\Gr_R(X)$,
\[
\Aut(\qcs{F}) \iso \Hom((\pi_0(X)_{\Fq})_{\Weil{}},\EEx),
\]
canonically.
\end{enumerate}
\end{theorem}

@@

\subsection{Group schemes that admit a Neron model}

We have just seen that if $X_K$ is a connected commutative algebraic group $X_K$ over a local field $K$ that admits a Neron model $X$ over $R$, then the category of   quasicharacter sheaves on the Greenberg transform of the Neron model $X$ provides a geometrization and categorification of quasicharacters of $X_K(K)$.

It is known that all Abelian varieties over local fields and all algebraic tori over local fields admit Neron models. In those two cases, we can improve Theorem~\ref{thm:QCXK} slightly with the following addendum.

\begin{proposition}\label{prop:natural_transformation}
Let $K$ be a local field with ring of integers $R$ and residue field $\Fq$.
Let $\mathcal{N}/K$ be the additive category of Abelian varieties over $K$ or the  additive category of algebraic tori over $K$.
Then the trace of Frobenius provides a natural transformation between the additive functors
\begin{enumerate}
\item[$F_{\QC}$:] $X_K \mapsto \QCiso{\Gr_R(X)}$
\item[$F_{\text{qc}}$:] $X_K \mapsto \Hom_{\text{qc}}(X_K(K),\EEx)$
\end{enumerate}
from the additive category $\mathcal{N}/K$ to the category of abelian groups.
\end{proposition}

It seems likely that there connected commutative algebraic groups over local fields that admit Neron models but are neither algebraic tori nor Abelian varieties; indeed, it is conjectured that K-wound unipotent group schemes over local fields like $\Fq((t))$ are examples. We have use the ponderous phrase "let $X_K$ be a connected commutative algebraic group over $K$ that admits a Néron model $X$" in Section~\ref{sec:applications} to draw attention to the fact that our theory of quasicharacter sheaves is applicable to more than algebraic tori and Abelian varieties.

@@

Let $K$ and $K'$ be local fields with rings of integers $R$ and $R'$ and residue fields $k$ and $k'$, respectively. Set $S = \Spec{S}$ and $S_n = \Spec{R_n}$; likewise define $S' = \Spec{R'}$ and $S'_{n'} = \Spec{R'_{n'}}$.

Let $X_K$ and $X'_{K'}$ be connected commutative algebraic groups over $K$ and $K'$, respectively, that admit Néron models $X$ and $X'$ over $R$ and $R'$, respectively.
Suppose $m$ and $m'$ are non-negative integers such that 
\begin{equation}\label{eq:schematic_transfer}
R_m \iso R'_{m'}
\qquad
\text{and}
\qquad
X \times_{S} S_m \iso X' \times_{S'} S'_{m'}.
\end{equation}
as smooth group schemes over $R_m$. 
Then
$
\Gr^R_m(X) \iso \Gr^{R'}_{m'}(X')
$
as smooth group schemes over $\Fq$; accordingly, by Proposition~\ref{prop:pullback} the isomorphism above determines an equivalence of categories
\begin{equation}\label{eq:categorical_transfer}
\QC(\Gr^R_m(X)) \cong \QC(\Gr^{R'}_{m'}(X'))
\end{equation}
which induces a group isomorphism
$
\QCiso{\Gr^R_m(X)} \cong \QCiso{\Gr^{R'}_{m'}(X')}
$
and thence, by trace of Frobenius, an isomorphism
\begin{equation}\label{eq:transfer}
\Hom_m(X_K(K),\EEx) = \Hom(X(R_m),\EEx) \iso \Hom(X'(R'_{m'}),\EEx) = \Hom_{m'}(X'_{K'}(K'),\EEx).
\end{equation}
Recall that $\Hom_m(X_K(K),\EEx)$ is a subgroup of $\Hom_\text{qc}(X_K(K),\EEx)$ and $\Hom_{m'}(X'_{K'}(K'),\EEx)$ is a subgroup of $\Hom_\text{qc}(X_K(K),\EEx)$.
The isomorphism \eqref{eq:transfer} is an instance of {\it transfer} of certain quasicharacters between $X_K(K)$ and $X_{K'}(K')$. 
When it exists, we now recognize this transfer of quasicharacters as a consequence of the equivalence of categories of quasicharacter sheaves \eqref{eq:categorical_transfer}.

Such a relationship can exist between quasicharacters of algebraic tori over local fields, even when the characteristics of $K$ and $K'$ differ.
Suppose $T$ and $T'$ are tori over non-archimedian local fields $K$ and $K'$,
splitting over $L$ and $L'$ respectively.
Then $T$ and $T'$ are said to be \emph{$N$-congruent} \cite{chai-yu:01a}*{\S 2} if there are isomorphisms
 \begin{align*}
  \alpha : \OL/\pi_K^N \OL &\to \OO{L'}/\pi_{K'}^N \OO{L'} \\
  \beta : \Gal(L/K) &\to \Gal(L'/K') \\
  \phi : X^*(T) &\to X^*(T')
 \end{align*}
 satisfying the conditions
 \begin{enumerate}
  \item $\alpha$ induces an isomorphism $\OK/\pi_K^N \OK \to \OO{K'}/\pi_{K'}^N \OO{K'}$,
  \item $\alpha$ is $\Gal(L/K)$-equivariant relative to $\beta$, and
  \item $\phi$ is $\Gal(L/K)$-equivariant relative to $\beta$.
 \end{enumerate}
If $T$ and $T'$ are $N$-congruent then $\alpha$, $\beta$ and $\phi$ determine an isomorphism 
\begin{equation}\label{transfer}
  \Hom_N(T(K), \EEx) \iso \Hom_N(T'(K'),\EEx).
\end{equation}
Note that if $T$ and $T'$ are $N$-congruent, then they are $N'$ congruent for every
$N' \leq N$, the isomorphism \eqref{transfer} is level-preserving at integers less than $N$.
%
One of the main results of \cite{chai-yu:01a} gives an isomorphism of group schemes between the
$\underline{T} \times_S S_n$ and $\underline{T}' \times_{S'} {S'}_n$ assuming that $T$ and $T'$ are sufficiently congruent.
They define a quantity $h$ as the smallest integer so that $\pi^h$ lies in the
Jacobian ideal associated to a natural embedding of $T$ into an induced torus \cite{chai-yu:01a}*{\S 8.1}.
Then it follows from \cite{chai-yu:01a}*{Thm. 8.5} that if $N > 3h$ and $T$ and $T'$ are $N$-congruent
then there is a canonical isomorphism of smooth group schemes
 $
  \TT_{N-3h} \to \TT'_{N-3h}
 $
 determined by $\alpha, \beta$ and $\phi$.
Combining this with the paragraph above gives an instance of the geometrization and categorification of the transfer of characters.

\begin{proposition}\label{prop:transfer}
 With notation above, suppose that $T$ and $T'$ are $N$-congruent and $N > 3h$.  Set $m \ceq N-3h$.
 Then there is a canonical equivalence of categories
 \[
 \QC(\Gr^{R}_m(\underline{T})) \to \QC(\Gr^{R'}_m(\underline{T}'))
 \]
 determined by $\alpha, \beta$ and $\phi$ inducing an isomorphism
 \[
\Hom_m(T(K), \EEx) \to  \Hom_m(T'(K'), \EEx)
 \]
 through the trace of Frobenius,  compatible with \eqref{transfer}.% ($m < N$).
\end{proposition}
